TBD

### 2020-03-11 manuscript/manuscript_clean.tex
incorporate final minor edits to accepted version.

\documentclass[11pt,]{article}
\usepackage[]{mathpazo}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            colorlinks=true,
            linkcolor=black,
            citecolor=black,
            urlcolor=black,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{amnat}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{}
    \pretitle{\vspace{\droptitle}}
  \posttitle{}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

@@

\emph{Keywords}: adaptive landscape, ecological networks,
eco-evolutionary dynamics, natural selection, host-parasitoid, community
context, extinction.

@@

\emph{Abstract Word Count}: 251

@@

\emph{Total Word Count}: 5833

@@

The loss of biodiversity is altering the structure of ecological
networks; however, we are currently in a poor position to predict how
these altered communities will affect the evolutionary potential of
remaining populations. Theory on adaptive landscapes provides a
framework for predicting how selection constrains phenotypic evolution,
but often treats the network context of evolving populations as a
``black box''. Here, we integrate ecological networks and adaptive
landscapes to examine how changes in food-web structure shape
evolutionary constraints. We conducted a field experiment that removed a
guild of larval parasitoids that imposed direct and indirect selection
pressures on an insect herbivore. We then measured herbivore survival as
a function of three key phenotypic traits to estimate directional,
quadratic, and correlational selection gradients in each treatment. We
used these selection gradients to characterize the slope and curvature
of the adaptive landscape to understand the direct and indirect effects
of consumer loss on evolutionary constraints. We found that the number
of traits under directional selection increased with the removal of
larval parasitoids, indicating greater selective constraints on the
trajectory of evolutionary change. Similarly, we found that the removal
of larval parasitoids altered the curvature of the adaptive landscape in
such a way that tended to decrease the evolvability of the traits we
measured in the next generation. Our results suggest that the loss of
trophic interactions can impose greater constraints on phenotypic
evolution. This indicates that the simplification of ecological
communities may constrain the adaptive potential of remaining
populations to future environmental change.

@@

The adaptive landscape provides a powerful framework for understanding
how natural selection has shaped the evolution of biodiversity ---from
genes to phenotypes to species
\citep{Wright1931, Simpson1944, Arnold2001}. More than a metaphor, the
adaptive landscape links quantitative genetic and phenotypic variation
to evolution by natural selection
\citep{Lande1979, Arnold1984applications, Arnold1984theory}, giving
insight to the direct and indirect effects of selective constraints
\citep{Arnold1992}. Ecological interactions often play a key role in
shaping selective constraints, as evidenced by the role of antagonistic
and mutualistic interactions in driving evolutionary change
\citep{Schluter2000, Abrams2000, Bronstein2006}. Although there is clear
evidence that species interactions can shape the adaptive landscape, we
still have a poor understanding of how the adaptive landscape is shaped
by a community of interacting species \citep{McPeek2017, terHorst2018}.
Resolution on how change in ecological communities shape phenotypic
evolution is urgently needed though, given the rapid losses of
biodiversity we are observing in the Anthropocene \citep{Scheffers2016}.

Ecological networks, such as food webs describing who-eats-whom, provide
an explicit representation of the direct and indirect effects that
emerge in a community of interacting species
\citep{Bascompte2014, McCann2012}. Here, we integrate ecological
networks and adaptive landscapes to understand how ecological
communities constrain evolutionary change. The direct and indirect
effects of natural selection can be inferred by quantifying the slope
and curvature of the adaptive landscape \citep{Arnold1992}. For example,
the slope is determined by directional selection gradients acting on
each phenotypic trait and influences the trajectory of evolutionary
change \citep{Lande1979, Arnold1992}. Selective constraints on evolution
increase with the number of traits under directional selection, as this
diminishes the number of optimal solutions \citep{Arnold2003}. The
curvature of the adaptive landscape is governed by nonlinear selection
gradients and acts to indirectly constrain evolution through its effect
on genetic constraints \citep{Arnold1992, Hansen2008}. Genetic
constraints are largely governed by a population's G-matrix ---the
additive genetic variances and covariances between traits
\citep{Hansen2008}. Stabilizing selection acts to erode genetic variance
in traits, which can impose a constraint on the ability of those traits
to respond to future selection \citep{Hansen2008}. Correlational
selection may alter the genetic covariance between traits, which may
facilitate or hinder future adaptation depending on the pattern of
selection on those traits \citep{Hansen2008}. If we want to predict how
ecological communities shape phenotypic evolution, we must understand
how ecological networks shape the adaptive landscape.

The loss of biodiversity is altering the structure of ecological
networks, which may influence evolutionary constraints in a number of
ways. For example, in a food web, if different consumers impose
directional selection on different traits of a shared resource, then a
more diverse consumer community may constrain evolution by increasing
the number of traits under selection. Alternatively, if consumers impose
selection on different values of a trait, then their selective effects
would cancel each other out in more diverse communities. To examine
these different possibilities, we conducted a field experiment that
removed a consumer guild that directly and indirectly impacts fitness of
an abundant insect herbivore (\emph{Iteomyia salicisverruca}, Family
Cecidomyiidae; fig. \ref{fig:Conceptual}). The larvae of this herbivore
induce tooth-shaped galls when they feed on the developing leaves of
willow trees \citep[\emph{Salix} sp.,][]{Russo2006}. These galls protect
larva from attack by generalist predators (e.g.~ants, spiders), but they
suffer high mortality from egg and larval parasitoids
\citep{Barbour2016}. We manipulated food-web structure by either
removing larval parasitoids (removal food web) or allowing both egg and
larval parasitoids to impose selection on gall midge traits (original
food web, fig. \ref{fig:Conceptual}). The relative abundance of these
larval parasitoids is much lower than the egg parasitoid
\citep{Barbour2016}, and thus represents a plausible extinction scenario
for this system. Larval parasitoids also impose indirect effects on gall
midge fitness through intraguild predation on the egg parasitoid (fig.
\ref{fig:Conceptual}). We applied modern statistical methods to quantify
how changes in food-web structure altered the slope and curvature of the
gall midge's adaptive landscape. Taken together, our study gives insight
to how the loss of biodiversity may constrain the evolution of
interacting populations.

\begin{figure}
\centering
\includegraphics{../analyses/complex_simple_foodwebs_v4.jpeg}
\caption{\label{fig:Conceptual}Experimental manipulation of food-web
structure associated with a leaf-galling midge (\emph{Iteomyia
salicisverruca}) feeding on the willow \emph{Salix hookeriana}. Black
arrows denote the flow of energy in this network of trophic
interactions. In the original food web, we allowed the full suite of egg
and larval parasitoids to impose selection. In the removal food web, we
used mesh bags to exclude the guild of larval parasitoids, only allowing
the egg parasitoid (\emph{Platygaster} sp.) to impose selection. Note
that larval parasitoids also impose indirect effects on gall midge
fitness through intraguild predation on the egg parasitoid. Larval
parasitoids include the following species (from left to right):
\emph{Mesopolobus} sp. (Family: Pteromalidae); \emph{Tetrastichus} sp.
(Family: Eulophidae); and \emph{Torymus} sp. (Family: Torymidae).}
\end{figure}

@@

We setup our food-web manipulation across 128 plants soon after galls
began developing on willows in early June of 2013. These 128 plants came
from eight different plant genotypes that spanned the range of trait
variation observed in this willow population \citep{Barbour2015}. For
the original food web (eight replicates per genotype), we used flagging
tape to mark 14 galled leaves per plant (\textasciitilde{}30 larvae),
allowing the full suite of egg and larval parasitoids to impose
selection. Marking galls with flagging tape ensured that all galls had
similar phenology when we collected galls later in the season. For the
removal food web, we enclosed 14 galled leaves with 10x15cm organza bags
(ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species
that attack during larval development. This treatment did not exclude
the egg parasitoid \emph{Platygaster} sp., which attacks prior to gall
initiation \citep[larva initiate gall development in Cecidomyiid
midges:][]{Gagne1989}. It was not possible to reciprocally manipulate
parasitoid attack (i.e.~exclude egg parasitoid, but allow larval
parasitoids), because it was not possible to identify midge oviposition
sites prior to gall formation. In late August, we collected marked and
bagged galls from each plant, placed them into 30 mL vials and kept them
in the lab for 4 months at room temperature. We then opened galls under
a dissecting scope and determined whether larvae survived to pupation
(our measure of fitness) or were parasitized. We did not include other
sources of mortality in our analyses, such as early larval death, as
they could influence the expression of the gall phenotype and confound
estimates of selection. For the food-web treatment that excluded larval
parasitoids, we further restricted our data by removing any incidental
instances of parasitism by a larval parasitoid. This represented less
than 3\% of the observations in this food-web treatment and allowed us
to focus our inferences of selection on those imposed by the egg
parasitoid. Our final dataset contains survival estimates for 1285
larvae from 613 galls and 111 plants.

@@

We collected data on three different traits that we expected to
influence larval survival based on previous work in this system
\citep{Barbour2016} and other work with gall midges
\citep{Weis1983, Heath2018}. First, we measured gall diameter as the
size of each gall chamber to the nearest 0.01 mm at its maximum diameter
(perpendicular to the direction of plant tissue growth). Previous work
in this system has shown that larger galls are associated with higher
survival \citep{Barbour2016}. Second, we measured clutch size by
counting the number of chambers in each gall
\citep{Weis1983, Heath2018}. All larvae collected from the same
multi-chambered gall were scored with the same clutch size. Third, we
measured oviposition preference as the density of larvae observed on a
plant in an independent survey. We did this by randomly sampling five
branches per tree and counting the number of individual gall chambers
(number of larvae). We then converted these counts to a measure of
larval density per 100 shoots by counting the number of shoots on the
last branch we sampled. All larvae collected from the same plant were
scored with the same oviposition preference. Measuring larval densities
on plants in the field is a common method for measuring oviposition
preference \citep{Gripenberg2010}; however, caution must be taken in
inferring `preference' as larval densities can be influenced by
processes other than preference \citep{Singer1986}. Fortunately, a
couple features of our study system suggest that larval densities may be
a good proxy for oviposition preference. For example, since our data
comes from a randomized placement of plant genotypes in a common garden,
there is no consistent bias in which plant genotypes females are exposed
to while searching for oviposition sites. Also, egg predation is a minor
source of mortality for galling insects in general \citep{Hawkins1997};
therefore, we do not expect any prior egg predation to bias our
estimates of observed larval densities.

\subsection{Quantifying the Adaptive
Landscape}\label{quantifying-the-adaptive-landscape}

Inferring adaptive landscapes assumes that trait distributions are
multivariate normal \citep{Lande1983}. To approximate this assumption,
we log-transformed clutch size and square-root transformed oviposition
preference. We then scaled all phenotypic traits (mean=0 and SD=1)
across treatments prior to our analyses (detailed below) to ensure that
our estimates of selection were comparable across traits and with other
studies.

Our analysis consisted of four parts. First, we used generalized linear
mixed models (GLMM) to quantify selection surfaces \textemdash linear
and nonlinear relationships between absolute fitness (\(W\)) and
standardized phenotypic traits (\(i\)) of individuals \textemdash in
each food-web treatment. Second, we translated selection surfaces into
the scale of relative fitness (\(w\)) in order to calculate standardized
selection gradients. Third, we used our estimates of directional
selection gradients to characterize the slope of the adaptive landscape,
which we used to measure selective constraints. Finally, we estimated
the curvature of the adaptive landscape and used a simulation to explore
its effects on the adaptive potential of the gall midge population in
the next generation.

\textbf{Selection surfaces}: Since larval survival was our measure of
absolute fitness, we used a GLMM that assumed a binomial error
distribution (and logit-link function). To approximate the selection
surface, we modelled larval survival as a function of food-web treatment
as well as linear (\(\alpha_i\)), quadratic (\(\alpha_{ii}\)), and
linear interactions (\(\alpha_{ij}\)) between each trait. We also
allowed these selection surfaces (\(\alpha\)) to vary between food-web
treatments. Note that to obtain valid estimates of linear selection
surfaces, we removed nonlinear terms prior to estimating linear
relationships \citep{Lande1983}. Other approaches have been advocated
for approximating selection surfaces \citep{Schluter1988}; however, our
approach enables us to calculate selection gradients, and thus is more
appropriate for approximating the adaptive landscape \citep{Arnold2003}.
To account for the nonindependence of clutch size (measured at gall
level) and oviposition preference (measured at plant level) as well as
any independent effects of willow genotype on larval survival, we
modelled gall ID nested within plant ID nested within genotype ID as
random effects. Although statistical models with random effects are not
common in analyses of natural selection, we think that modelling random
effects can mitigate biased estimates of selection due to environmental
covariances between traits and fitness \citep{Rausher1992}. Since our
end goal was to characterize the relationship between mean trait values
and mean fitness (adaptive landscape), we assumed the mean value of our
random effects (i.e.~setting them to zero) when calculating selection
surfaces. We then used parametric bootstrapping (1,000 replicates) to
estimate the effect of food-web treatment on larval survival as well as
selection surfaces in each food-web treatment. To determine whether
trait-fitness relationships differed between food-web treatments, we
calculated the difference in bootstrapped replicates between treatments.

\textbf{Selection gradients}: Selection gradients cannot be estimated
directly from GLMMs of selection surfaces because the response is in
terms of absolute fitness and the coefficients are on a nonlinear scale.
For example, the coefficients in the above model measure the change in
the logarithm of the odds of surviving (i.e.
\(\text{ln}\{W(z)/[1-W(z)]\}\)) associated with 1 SD change in a trait
with all other traits held fixed at their mean. Therefore, we used the
method developed by \citet{Janzen1998} to translate selection surfaces
from the above model into the scale of relative fitness in order to
calculate directional (\(\beta_i\)), quadratic (\(\gamma_{ii}\)), and
correlational (\(\gamma_{ij}\)) selection gradients. Briefly, this
method calculates the average gradient of selection surfaces by
multiplying the average of \(W(z)[1-W(z)]\) by each regression
coefficient (e.g. \(\alpha_i\), \(\alpha_{ii}\), or \(\alpha_{ij}\)). We
then divided this average gradient by the mean fitness (\(\bar W\)) to
put it on the scale of relative fitness, and thus interpretable as a
selection gradient. We estimated selection gradients separately for each
food-web treatment. We also determined whether selection gradients
differed between food-web treatments by calculating the difference in
bootstrapped replicates between treatments. Note that we doubled all
quadratic terms prior to calculating selection gradients to put them on
the same scale as estimates of directional and correlational selection
\citep{Stinchcombe2008}.

\textbf{Selective constraints}: The number of selective constraints is
determined by the slope of the adaptive landscape, which in our study
corresponds to:

>>>>

\documentclass[11pt,]{article}
\usepackage[]{mathpazo}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{
            colorlinks=true,
            linkcolor=black,
            citecolor=black,
            urlcolor=black,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{amnat}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

\author{}
\date{\vspace{-2.5em}}

@@

\emph{Keywords}: fitness landscape, adaptive landscape, ecological
networks, eco-evolutionary dynamics, natural selection, host-parasitoid,
community context, extinction.

@@

\emph{Abstract Word Count}: 256

@@

\emph{Total Word Count}: 6119

@@

The loss of biodiversity is altering the structure of ecological
networks; however, we are currently in a poor position to predict how
these altered communities will affect the evolutionary potential of
remaining populations. Theory on fitness landscapes provides a framework
for predicting how selection alters the evolutionary trajectory and
adaptive potential of populations, but often treats the network of
interacting populations as a ``black box''. Here, we integrate
ecological networks and fitness landscapes to examine how changes in
food-web structure shape phenotypic evolution. We conducted a field
experiment that removed a guild of larval parasitoids that imposed
direct and indirect selection pressures on an insect herbivore. We then
measured herbivore survival as a function of three key phenotypic traits
to estimate directional, quadratic, and correlational selection
gradients in each treatment. We used these selection gradients to
characterize the slope and curvature of the fitness landscape to
understand the direct and indirect effects of consumer loss on
phenotypic evolution. We found that the number of traits under
directional selection increased with the removal of larval parasitoids,
indicating evolution was more constrained toward a specific combination
of traits. Similarly, we found that the removal of larval parasitoids
altered the curvature of the fitness landscape in such a way that tended
to decrease the evolvability of the traits we measured in the next
generation. Our results suggest that the loss of trophic interactions
can impose greater constraints on phenotypic evolution. This indicates
that the simplification of ecological communities may constrain the
adaptive potential of remaining populations to future environmental
change.

@@

The fitness landscape provides a powerful framework for understanding
how natural selection has shaped the evolution of biodiversity ---from
genes to phenotypes to species
\citep{Wright1931, Simpson1944, Arnold2001}. More than a metaphor, the
fitness landscape links quantitative genetic and phenotypic variation in
multiple traits to evolution by natural selection
\citep{Lande1979, Arnold1984applications, Arnold1984theory}. Ecological
interactions often play a key role in shaping natural selection, as
evidenced by the role of antagonistic and mutualistic interactions in
driving evolutionary change
\citep{Schluter2000, Abrams2000, Bronstein2006}. Although there is clear
evidence that species interactions can shape the fitness landscape, we
still have a poor understanding of how the fitness landscape is shaped
by a community of interacting species
\citep{McPeek2017, terHorst2018, Hui2018}. Resolution on how change in
ecological communities shape phenotypic evolution is urgently needed
though, given the rapid losses of biodiversity we are observing in the
Anthropocene \citep{Scheffers2016}.

Ecological networks, such as food webs describing who-eats-whom, provide
an explicit representation of the direct and indirect effects that
emerge in a community of interacting species
\citep{Bascompte2014, McCann2012}. Here, we integrate ecological
networks and fitness landscapes to understand how selection imposed by
ecological communities alter the evolutionary trajectory and adaptive
potential of interacting populations \citep{Hui2018}. The effects of
natural selection on multiple phenotypic traits can be inferred by
quantifying the slope and curvature of the fitness landscape
\citep{Arnold1992}. The slope is determined by the strength of
directional selection acting on each trait, which influences the
trajectory of evolution \citep{Lande1979, Arnold1992}. This fact is made
clear by the `Lande equation', \(\Delta\bar z = \text{G}\beta\), where
\(\Delta \bar z\) is a column vector of the average change in each trait
between generations, \text{G} is the additive genetic (co)variance
matrix for these traits (i.e.~G-matrix), and \(\beta\) is a column
vector of directional selection gradients acting on each trait
(i.e.~slope). Grant and Grant \citeyearpar{Grant1995} used the slope of
the fitness landscape for multiple traits related to body and beak size
in Darwin's finches to accurately predict the effects of drought on the
evolutionary trajectory of these traits. The curvature of the fitness
landscape is governed by the strength of stabilizing, disruptive, and
correlational selection acting on each trait, which can alter the
evolutionary potential of a population through its effect on the
G-matrix \citep{Hansen2008}. For example, stabilizing selection acts to
erode genetic variance in a trait, which can impose a constraint on the
ability of this trait to respond to future selection \citep{Hansen2008}.
In contrast, disruptive selection toward extreme trait values acts to
increase genetic variance in a trait, thus increasing the capacity for
future adaptation \citep{Bolnick2008}. Correlational selection alters
the genetic covariance between traits, which may facilitate or hinder
future adaptation depending on the pattern of selection on those traits
and the structure of the G-matrix \citep{Brodie1992}. If we want to
predict how ecological communities shape phenotypic evolution, we must
understand how ecological networks shape the fitness landscape of
interacting populations.

The loss of biodiversity is altering the structure of ecological
networks \citep{Landi2018ComplexityStability}, which may influence the
slope and curvature of the fitness landscape in a number of ways. For
example, consider how changes in consumer diversity in a food web may
alter the slope of the fitness landscape for a shared resource. If
different consumers impose directional selection on different traits of
the resource, then a more diverse consumer community would increase the
number of traits under selection, which may constrain the trajectory of
evolution toward a specific phenotype. Alternatively, if consumers
impose selection on different values of a trait, then their selective
effects would cancel each other out in more diverse communities, which
would allow for a greater diversity of phenotypes to persist. Now
consider the effects of consumer diversity on the curvature of the
fitness landscape. If consumers impose selection on different ends of a
resource's trait distribution, then a more diverse consumer community
may impose stabilizing selection, which would decrease genetic variance
in this trait. In contrast, additional consumers may impose disruptive
selection if their cumulative effect decreases the relative fitness of a
resource's average trait value, which would increase the genetic
variance in this trait.

To examine these different possibilities, we conducted a field
experiment that removed a consumer guild that parasitizes an abundant
insect herbivore (\emph{Iteomyia salicisverruca}, Family Cecidomyiidae;
fig. \ref{fig:Conceptual}). The larvae of this herbivore induce
tooth-shaped galls when they feed on the developing leaves of willow
trees \citep[\emph{Salix} sp.,][]{Russo2006}. These galls protect larva
from attack by generalist predators (e.g.~ants, spiders), but they
suffer high mortality from egg and larval parasitoids
\citep{Barbour2016}. We manipulated food-web structure by either
removing larval parasitoids (removal food web) or allowing both egg and
larval parasitoids to impose selection on gall midge traits (original
food web, fig. \ref{fig:Conceptual}). Larval parasitoids also impose
indirect effects on gall midge fitness through intraguild predation on
the egg parasitoid (fig. \ref{fig:Conceptual}). We applied modern
statistical methods to quantify how changes in food-web structure
altered the slope and curvature of the gall midge's fitness landscape.
Taken together, our study gives insight to how the loss of biodiversity
may alter both the evolutionary trajectory and adaptive potential of
interacting populations.

\begin{figure}
\centering
\includegraphics{../analyses/complex_simple_foodwebs_v4_5.jpg}
\caption{\label{fig:Conceptual}Experimental manipulation of food-web
structure associated with a leaf-galling midge (b, \emph{Iteomyia
salicisverruca}) feeding on the willow \emph{Salix hookeriana} (a).
Black arrows denote the flow of energy in this network of trophic
interactions. In the original food web, we allowed the full suite of egg
and larval parasitoids to impose selection. In the removal food web, we
used mesh bags to exclude the guild of larval parasitoids, only allowing
the egg parasitoid (c, \emph{Platygaster} sp.) to impose selection. Note
that larval parasitoids also impose indirect effects on gall midge
fitness through intraguild predation on the egg parasitoid. Larval
parasitoids include the following species: \emph{Mesopolobus} sp. (d,
Family: Pteromalidae); \emph{Tetrastichus} sp. (e, Family: Eulophidae);
and \emph{Torymus} sp. (f, Family: Torymidae).}
\end{figure}

@@

We setup our food-web manipulation across 128 plants soon after galls
began developing on willows in early June of 2013. These 128 plants came
from 8 different plant genotypes that spanned the range of trait
variation observed in this willow population \citep{Barbour2015}. For
the original food web (eight replicates per genotype), we used flagging
tape to mark 14 galled leaves per plant (\textasciitilde{}30 larvae),
allowing the full suite of egg and larval parasitoids to impose
selection. Marking galls with flagging tape ensured that we compared
galls with similar phenology in both treatments when we collected galls
later in the season. For the removal food web, we enclosed 14 galled
leaves with 10x15cm organza bags (ULINE, Pleasant Prairie, WI, USA) to
exclude three parasitoid species that attack during larval development.
This treatment did not exclude the egg parasitoid \emph{Platygaster}
sp., which attacks prior to gall initiation \citep[larva initiate gall
development in Cecidomyiid midges:][]{Gagne1989}. It was not possible to
reciprocally manipulate parasitoid attack (i.e.~exclude egg parasitoid,
but allow larval parasitoids), because it was not possible to identify
midge oviposition sites prior to gall formation. In late August, we
collected marked and bagged galls from each plant, placed them into 30
mL vials and kept them in the lab for 4 months at room temperature. We
then opened galls under a dissecting scope and determined whether larvae
survived to pupation (our measure of fitness) or died due to parasitism.
We further distinguished whether mortality was caused by an egg or
larval parasitoid. Early larval death was another important source of
mortality (17\%), but we excluded these cases from our analysis. We did
this because early larval death can stunt gall growth, which would
confound estimates of selection due to parasitism on one of the
phenotypes we measured (chamber diameter; details given in next
section). For the food-web treatment that excluded larval parasitoids,
we further restricted our data by removing any incidental instances of
parasitism by a larval parasitoid. This represented less than 3\% of the
observations in this food-web treatment and allowed us to focus our
inferences of selection on those imposed by the egg parasitoid. Our
final dataset contains survival estimates for 1285 larvae from 613 galls
and 111 plants.

@@

We collected data on three different traits that we expected to
influence larval survival based on previous work in this system
\citep{Barbour2016} and other work with gall midges
\citep{Weis1983, Heath2018}. First, we measured gall diameter as the
size of each gall chamber to the nearest 0.01 mm at its maximum diameter
(perpendicular to the direction of plant tissue growth). Previous work
in this system has shown that larger galls are associated with higher
survival \citep{Barbour2016}. Second, we measured clutch size by
counting the number of chambers in each gall
\citep{Weis1983, Heath2018}. All larvae collected from the same
multi-chambered gall were scored with the same clutch size. Third, we
measured oviposition preference for individual plants as the density of
larvae observed on a plant in an independent survey. We did this by
randomly sampling five branches per tree and counting the number of
individual gall chambers (number of larvae). We then converted these
counts to a measure of larval density per 100 shoots by counting the
number of shoots on the last branch we sampled. All larvae collected
from the same plant were scored with the same oviposition preference.
Measuring larval densities on plants in the field is a common method for
measuring oviposition preference \citep{Gripenberg2010}; however,
caution must be taken in inferring `preference' as larval densities can
be influenced by processes other than preference \citep{Singer1986}.
Fortunately, two features of our study system suggest that larval
densities may be a good proxy for oviposition preference. First, since
our data comes from a randomized placement of plant genotypes in a
common garden, females midges have equal exposure to many possible plant
genotypes when choosing an oviposition site. Second, egg predation is a
minor source of mortality for galling insects in general
\citep{Hawkins1997}; therefore, we do not expect any prior egg predation
to bias our estimates of observed larval densities.

\subsection{Quantifying the Fitness
Landscape}\label{quantifying-the-fitness-landscape}

Inferring the fitness landscape assumes that trait distributions are
multivariate normal \citep{Lande1983}. To approximate this assumption,
we log-transformed clutch size and square-root transformed oviposition
preference. Chamber diameter already had a normal distribution so we did
not transform it. We then scaled all phenotypic traits (mean=0 and SD=1)
across treatments prior to our analyses (detailed below) to ensure that
our estimates of selection were comparable across traits and with other
studies.

Our analysis consisted of four parts. First, we used generalized linear
mixed models (GLMM) to quantify selection surfaces \textemdash linear
and nonlinear relationships between absolute fitness (\(W\)) and
standardized phenotypic traits (\(i\)) of individuals \textemdash in
each food-web treatment. Second, we scaled selection surfaces relative
to mean fitness (\(\bar W\)) in order to calculate standardized
selection gradients. Third, we used our estimates of directional
selection gradients to characterize the slope of the fitness landscape,
which we used to quantify the effects of food-web structure on the
trajectory of evolution. Finally, we estimated the curvature of the
fitness landscape and used a simulation to explore its effects on the
adaptive potential of the gall midge population in the next generation.

\textbf{Selection surfaces}: Since larval survival was our measure of
absolute fitness, we used a GLMM that assumed a binomial error
distribution (and logit-link function). To approximate the selection
surface, we modelled larval survival as a function of food-web treatment
as well as linear (\(\alpha_i\)), quadratic (\(\alpha_{ii}\)), and
statistical interactions (\(\alpha_{ij}\)) between each trait. Note that
to obtain valid estimates of linear selection surfaces, we removed
nonlinear terms prior to estimating linear relationships
\citep{Lande1983}. Other approaches have been advocated for
approximating selection surfaces \citep{Schluter1988}; however, our
approach enables us to calculate selection gradients, and thus is more
appropriate for approximating the fitness landscape \citep{Arnold2003}.
To account for the nonindependence of clutch size (measured at gall
level) and oviposition preference (measured at plant level) as well as
any independent effects of willow genotype on larval survival, we
modelled gall identity nested within plant identity nested within
genotype identity as random effects. Although statistical models with
random effects are not common in analyses of natural selection, we think
that modelling random effects can mitigate biased estimates of selection
due to environmental covariances between traits and fitness
\citep{Rausher1992}. Since our end goal was to characterize the
relationship between mean trait values and mean fitness (fitness
landscape), we assumed the mean value of our random effects
(i.e.~setting them to zero) when calculating selection surfaces. We then
used parametric bootstrapping (1,000 replicates) to estimate the effect
of food-web treatment on larval survival as well as selection surfaces
in each food-web treatment. To determine whether trait-fitness
relationships differed between food-web treatments, we calculated the
difference in bootstrapped replicates between treatments.

\textbf{Selection gradients}: Selection gradients cannot be estimated
directly from GLMMs of selection surfaces because the response is in
terms of absolute fitness and the coefficients are on a nonlinear scale.
For example, the coefficients in the previously described binomial model
measure the change in the log-odds of survival associated with 1 SD
change in a trait with all other traits held fixed at their mean.
Therefore, we used the method developed by \citet{Janzen1998} to
translate selection surfaces from the above model into the scale of
relative fitness in order to calculate directional (\(\beta_i\)),
quadratic (\(\gamma_{ii}\)), and correlational (\(\gamma_{ij}\))
selection gradients. Briefly, this method calculates the average
gradient of selection surfaces by multiplying the average of
\(W(z)[1-W(z)]\) by each regression coefficient (e.g. \(\alpha_i\),
\(\alpha_{ii}\), or \(\alpha_{ij}\)). We then divided this average
gradient by the mean fitness (\(\bar W\)) to put it on the scale of
relative fitness (\(w=W/\bar W\)), and thus interpretable as a selection
gradient. We estimated selection gradients separately for each food-web
treatment. We also determined whether selection gradients differed
between food-web treatments by calculating the difference in
bootstrapped replicates between treatments. Note that we doubled all
quadratic terms prior to calculating selection gradients to put them on
the same scale as estimates of directional and correlational selection
\citep{Stinchcombe2008}.

\textbf{Evolutionary trajectory}: The effect of selection on the
trajectory of evolution is determined by the slope of the fitness
landscape, which in our study corresponds to:

\[\textbf{Slope} = \begin{pmatrix} \beta_{\text{Diam}} \\ \beta_{\text{Clutch}} \\ \beta_{\text{Pref}} \end{pmatrix} \]
where each \(\beta_i\) corresponds to the directional selection gradient
acting on each trait. By comparing selection gradients in each food-web
treatment (i.e.~95\% CI does not overlap zero), we can infer the effect
of food-web structure on the trajectory of phenotypic evolution.
